---
author: "Ronit Rudra"
date: "Monday, June 08, 2015"
---

*PEER ASSESSMENT 1*
=================

#### This Assignment is a part of the course [Reproducible Research](https://www.coursera.org/course/repdata) provided by Johns Hopkins University on [Coursera](https://www.coursera.org/) ####

The dataset used in this assignment is: [Activity Monitoring Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)

This report analyzes the aforementioned dataset and answers the following questions:
<ul>
<li> `What is mean total number of steps taken per day?` </li>
<li> `What is the average daily activity pattern?` </li>
<li> `Input missing values.` </li>
<li> `Are there differences in activity patterns between weekdays and weekends?` </li>
</ul>
###The following describes the methodologies and procedures for answering the aforementioned questions:###

#### *STEP 1: Import Data* ####

1. The following R code snippet imports the data from the URL. Since the dataset is contained inside a zip archive, it needs to be extracted first.

```{r, echo=TRUE,cache=TRUE}
# Create a temporary directory
temp <- tempfile()
# Download file to the temp directory
download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
# Unzip the file and read the CSV file
activity <- read.csv(unz(temp, "activity.csv"),stringsAsFactors=FALSE)
#unlink the temp directory
unlink(temp)
#View summary of the dataset
summary(activity)
#View the structure of the dataset
str(activity)
```

#### *STEP 2: Transform data* ####

1. From the above step, the structure of the `activity` dataset shows that one of the columns has dates. Hence, it is prudent to convert the character class of `date` from char to the Date format.

```{r, echo=TRUE}
# Convert char to Date
activity$date <- as.Date(activity$date)
str(activity)
```

2. Now, the dates are in the correct format and will enable easier manipulation.

#### *STEP 3: Analysis of activity per day (question 1)* ####

1. For calculating the parameters on a per day basis, we need to split the `activity` dataframe into sub data frames

```{r, echo=TRUE}
# Use the split function to split the data
by_date <- split(activity,activity$date)
# Calculate summed up value
summed <- data.frame(sapply(by_date,function(x){sum(x$steps,na.rm=TRUE)}))
print(summed)
```

2. The above method is a bit complicated and poses problems when plotting.

3. The second method is much easier and utilizes the `dplyr` library function `summarize`

```{r, echo=TRUE, message=FALSE}
# Load the dplyr library
library(dplyr)
#Create a data frame of mean values per day using summarise
summed <- summarize(group_by(activity,date),total_steps=sum(steps,na.rm=TRUE))
print(summed)
```

Note: the `naN` indicates the presence of all `NA` values for the particular day.

4. To plot the histogram of `total steps` we use both the `base` and `ggplot2` plotting system

```{r,echo=TRUE,message=FALSE,cache=TRUE}
#base system
hist(summed$total_steps,xlab="Total Steps",ylab="Count",col="steelblue",
     main="Histogram for total steps")
# ggplot2 system
library(ggplot2)
g<-ggplot(summed,aes(total_steps))
g+geom_histogram(stat="bin",fill="steelblue",color="black")+labs(x="Total Steps",y="Count",title="Histogram for total steps")
```

5. Plot of the `total_steps` versus `date`

```{r,echo=TRUE,message=FALSE,cache=TRUE}
#ggplot2 has already been loaded
# Define the base layer of the plot
sum <- ggplot(data=summed,aes(x=date,y=total_steps))
# Add the histogram layer
sum <- sum + geom_bar(stat="identity",fill="steelblue")
# Add labels
sum <- sum + labs(x="Date",y="Total Steps Taken",title="Plot of Total Steps versus Date")
# plot the graph
plot(sum)
```

6. Calculating the mean and median:

```{r,echo=TRUE,message=FALSE}
paste("The mean is:",round(mean(summed$total_steps),2))
paste("The median is:",round(median(summed$total_steps),2))
```

#### *STEP 4: Average Daily Activity Pattern* ####

1. For this we need to calculate the average number of steps taken per interval averaged over all of the 61 days.

```{r,echo=TRUE}
mean_interval <- summarize(group_by(activity,interval),Mean=mean(steps,na.rm=TRUE))
print(mean_interval)
```

2. Potting the Average steps versus the interval using `ggplot2`

```{r,echo=TRUE,cache=TRUE}
# Make 1st ggplot layer
act<-ggplot(mean_interval,aes(interval,Mean))
# Add points, lines and labels
act+geom_line(lwd=.5,color="red")+geom_point(color="orange",alpha=0.5)+
    labs(x="Interval",y="Average Steps",title="Average Number of Steps Taken (Averaged Across All Days)")
```

3. Finding the interval with highest average steps

```{r,echo=TRUE}
mean_interval[which.max(mean_interval$steps),]
```

#### *STEP 5: Input Missing Values* ####

1. The `activity` dataset contains a number of instances in the `steps` attribute which have missing values denoted by `NA`. The total number of missing values can be calculated as:

```{r,echo=TRUE}
sum(is.na(activity$steps))
```

2. To fill in the missing values we choose to assign them the mean value for that interval. The following code snippet inserts suitable values to the `NA` rows.

```{r,echo=TRUE}
# Define factor list to split the data on. Saved separately to ease unsplit
factor_interval <- factor(activity$interval)
# Split dataframe on factors based on the dates
splitted <- split(activity,factor_interval)
# Fill in NA values for each sub data frame using mean of the intervals
splitted <- lapply(splitted,function(x){
    indx <- which(is.na(x$steps))            # index number of rows with NA
    # mean_interval contains mean values of steps in each interval
    x[indx,1] <- mean_interval[x[indx,3]==mean_interval[,1],2]  # Replace "steps"       
    x
    }
    )
# Unsplit the data frame to a new dataframe as required
activity_no_NA <- unsplit(splitted,factor_interval)
print("The New Dataest with filled in NA values:")
head(activity_no_NA)
#check for missing values
sum(is.na(activity_no_NA))

```

3. Difference between `activity` and `activity_no_NA`. We plot the histogram for total steps taken using the `ggplot2` plotting system. We also calculate the mean and median for the new dataset.

```{r,echo=TRUE,cache=TRUE}
# Sum up steps taken each day
summed_no_NA <- summarize(group_by(activity_no_NA,date),total_steps=sum(steps))
# Plot histogram
act2<-ggplot(summed_no_NA,aes(total_steps))
act2<-act2 + geom_histogram(stat="bin",color="black",fill="steelblue") + 
    labs(x="Total Steps",y="Count",title="Histogram for total steps")
plot(act2)
# Calculate mean and median
paste("The mean is:",round(mean(summed_no_NA$total_steps),2))
paste("The median is:",round(median(summed_no_NA$total_steps),2))
```

4. It is obseverved that the `mean` value moves closer to the middle and the `median` is now closer to the mean than before.

#### *STEP 6: Activity patterns between Weekdays and Weekends* ####

1. First we need to classify the dates as Weekdays or Weekdends. For that purpose we utilize the `weekdays()` function which returns the day of the week based on date.

```{r,echo=TRUE}
factor_day<-weekdays(activity$date)     # Returns character vector with days
levels(factor(factor_day))
```

2. A list is created classifying the days into weekdays or weekends

```{r,echo=TRUE}
class_day<-sapply(factor_day,function(x){
    if(x=="Sunday" | x=="Saturday")
        y<-"Weekend"
    else
        y<-"Weekday"
    y
}
)
```

3. The list is added as a column to the `activity` dataset

```{r,echo=TRUE}
activity_no_NA$dayClass<-factor(class_day)
head(activity_no_NA)
```

4. The mean of steps in each interval on weekdays and weekends is calculated

```{r,echo=TRUE}
day_sum<-summarize(group_by(activity_no_NA,dayClass,interval),
                   mean_steps=mean(steps,rm.na=T))
print(day_sum)
```

5. The mean number of steps is plotted for both weekdays and weekends

```{r,echo=TRUE,cache=TRUE}
wkd_wkend <- ggplot(data=day_sum,aes(interval,mean_steps))
wkd_wkend + facet_grid(.~dayClass)+geom_line(lwd=0.5,color="orange")+
    labs(x="Intervals (5 mins)",y="Mean Steps",title="Activity Comparison between Weekdays and Weekends")
```

__*Hence all the questions have been suitably answered with relevant codes and plots*__
